Pattern Avoidance

### 2014-11-08 project2/paper/paper.tex
changed attributions
another edit on Section 4
Cleaned up some notation

In Sections \ref{312} and \ref{321} we consider sequences $s_n(\pi)$ for $\pi$ an element of  \\ $\{123,132,213,231,312,321\}$, and prove the following theorem.

\begin{theorem*}
For $\pi$ an element of $\{123,132,213,231,312,321\}$, the sequence $(s_n(\pi))$ is the Catalan numbers.
\end{theorem*}

Of particular interest is the fact that two entirely distinct methods of proof had to be used to prove this theorem for all six permutations. In Section \ref{S4} we will consider sequences $s_n(\pi)$ for $\pi$ a permutation of $\{1,2,3,4\}$ and state a conjecture concerning $(s_n(\pi))$. In Section \ref{Tn} we consider a variation of the avoidance problem that considers the problem of avoiding sequences in a subset $T_n$ of the permutations of $n$ elements. Before proceeding, however, we formally introduce the notation that will be used throughout this paper.

@@

\emph{Yajit wrote this section. Leon closely edited this section.}

@@

%\subsection{Generating Trees}
%
%The generating tree for the sequence $s_n(\pi)$ is the infinite tree whose vertices are elements of $S_n$ that avoid $\pi$. The first node in the tree is always 1. To construct the tree, we define the children of an arbitrary node. Let $\omega\in S_k$ be a node of the generating tree of $s_n(\pi)$. Then the children of $\omega$ are elements of $S_{k+1}$ that avoid $\pi$ and that can be obtained by inserting the integer $k+1$ between integers in $\omega$. Figure \ref{fig:M1} is an example of the generating tree of $s_n(312)$. 
%
%\begin{figure}[h!]
%\begin{tikzpicture}[level distance=0.5cm, growth parent anchor={south}, nodes={anchor=north}]
%\tikzstyle{hollow node}=[circle,draw,inner sep=0.1mm, text width=1.36cm,align=center]
%\tikzstyle{level 1}=[sibling distance=6cm] 
%\tikzstyle{level 2}=[sibling distance=3cm] 
%\tikzstyle{level 3}=[sibling distance=1.8cm]
%\tikzstyle{level 4}=[sibling distance=0.9cm]
%\node [hollow node]{1}
%	child {
%		node[hollow node] {12}
%		child{
%			node[hollow node]{123}
%			child{node[hollow node]{1234}}
%			child{node[hollow node]{1243}}
%		}
%		child{
%			node[hollow node]{132}
%			child{node{\vdots}}
%		}
%	}
%    	child {
%		node[hollow node] {21}
%    		child {
%			node[hollow node] {213}
%			child{node{\vdots}}
%		}
%      		child {
%			node[hollow node] {231}
%			child{node{\vdots}}
%		}
%		child {
%			node[hollow node] {321}
%			child{node{\vdots}}
%		}
%    	};
%
%\end{tikzpicture}
%\caption{Generating tree of the sequence $s_n(312)$} \label{fig:M1}
%\end{figure}
\subsection{Reversing}

\begin{lemma}[Reversing Lemma]
The permutation $\sigma$ avoids the permutation $\pi$ if and only if $\mathcal{R}(\sigma)$ avoids $\mathcal{R}(\pi)$. I.e. $\mathcal{R}$ is an avoidance preserving operator. 
\end{lemma}
\begin{proof}
We need to only prove one direction of this theorem, then the other direction follows because $\mathcal{R}^2=1$. We show that if ${\sigma}\in S_n$ avoids $\pi$ then $\mathcal{R}(\sigma)$ avoids $\mathcal{R}(\pi)$. Suppose that $\mathcal{R}(\sigma)$ does not avoid $\mathcal{R}(\pi)$. Then If we reverse $\mathcal{R}(\sigma)$ again, the subsequence of $\mathcal{R}(\sigma)$ that follows the same pattern as $\mathcal{R}(\pi)$ will reverse to a subsequence that follows the same pattern as $\pi$, giving us a contradiction. Therefore $\mathcal{R}(\sigma)$ avoids $\mathcal{R}(\pi)$.

@@

\begin{corollary} 
For a permutation $\pi$, $s_n(\pi)=s_n(\mathcal{R}(\pi))$. 
\end{corollary}
\begin{proof}
This result follows because the reversing lemma gives a bijection between permutations that avoid $\pi$ and permutations that avoid $\mathcal{R}(\pi)$. 
\end{proof}

\subsection{Flipping}
In this section we introduce the concept of flipping to prove the `Flipping Lemma.'
\begin{definition}
We define the \emph{flip} of a sequence $a$ as the sequence $b$ with the same elements as $a$, but with the largest element swapped with the smallest element, the second largest element swapped with the second smallest element, etc.  The flipping operator will be denoted by $\mathcal{F}$.  
\end{definition}

@@

\begin{lemma}[Flipping Lemma]
The permutation $\sigma$ avoids the permutation $\pi$ if and only if $\mathcal{F}(\sigma)$ avoids $\mathcal{F}(\pi)$. I.e. $\mathcal{F}$ is an avoidance preserving operator. 
\end{lemma}
\begin{proof}
Notice that we only need to prove one direction of this statement. The opposite direction will follow since applying $\mathcal{F}$ twice returns the original permutation. We show that if ${\sigma}\in S_n$ avoids $\pi$ then $\mathcal{F}(\sigma)$ avoids $\mathcal{F}(\pi)$. Suppose that $\mathcal{F}(\sigma)$ does not avoid $\mathcal{F}(\pi)$. Let $\mathcal{F}(\sigma)=a_1\cdots a_n$ and let $\mathcal{F}(\pi)=b_1\cdots  b_k$. Then there is some subsequence $a_{i_1}\cdots a_{i_k}$ so that $a_{i_c}<a_{i_d}$ if and only if $b_c<b_d$. Now flipping $\mathcal{F}(\sigma)$ gives us the sequence $(n+1-a_1)(n+1-a_2)\cdots(n+1-a_n)$ and flipping $\mathcal{F}(\pi)$ gives the sequence $(k+1-b_1)(k+1-b_2)\cdots(k+1-b_k)$. Thus $n+1-a_{i_c}>n+1-a_{i_d}$ if and only if $k+1-b_c>k+1-b_d$. So  this implies that $\sigma$ does not avoid $\pi$, a contradiction. 
\end{proof}

\begin{corollary} 
For a permutation $\pi$, $s_n(\pi)=s_n(\mathcal{F}(\pi))$. 
\end{corollary}
\begin{proof}
This result follows because the flipping lemma gives a bijection between permutations that avoid $\pi$ and permutations that avoid $\mathcal{F}(\pi)$. 
\end{proof}

@@

\emph{Deepak wrote this section. Leon closely edited this section.}

%It is easy to see that the above proof can be modified slightly to prove that the sequences $s_n(132)$, $s_n(213)$ and $s_n(231)$ are in fact the sequence of catalan numbers as well. In fact, we claim that if $\sigma$ and $\pi$ are both permutations in $S_k$ for some arbitrary $k$, then if $\sigma = flip(\pi)$, $s_n(\pi) = s_n(\sigma)$. This is easy to see since if $a$ is a permutation that avoids $\sigma$, then $flip(a)$ is a permutation that avoids $\pi$ -- since this can be done both ways, it is bijective and we can conclude that the cardinalities of the two sets are equal, or $s_n(\sigma) = s_n(\pi)$.\todo{we should move this up and make it clear from the beginning that this stuff generalizes}

@@

\label{321}
Unfortunately, there is no clear way to reformulate Lemma \ref{lemma1} for the case of avoiding the permutation 321 or 123. As a consequence, the proof in the previous section simply does not hold. Fortunately, however, we still have a closed form expression for $s_n(321)$ (recall that $s_n(321)=s_n(123)$ by our flipping lemma). Surprisingly, in fact, the sequence remains the same: $s_n(321)=C_n$, the $n^{th}$ Catalan number. In order to prove this result, we need the machinery of Young tableaux. What follows is a brief exposition of the key results, adapted from Stanley's \emph{Algebraic Combinatorics}.

@@

\emph{Leon wrote this section. Yajit edited this section.}

@@

Since $T_n$ is only defined for even $n$, henceforth we shall refer to $T_n$ as $T_{2m}$ where $m$ is any integer greater than or equal to $1$.

@@

Unfortunately, it turns out that proving the equivalence between $231$ and $213$ is not as straightforward as above. In fact, however, we can prove a stronger result: both $t_n(231)$ and $t_n(213)$ are equal to $2^{n-1}$, and hence the sequences we obtain are equivalent.

\begin{theorem}
$t_n(231)$, the number of permutations in $T_n$ avoiding the pattern 231, is equal to $2^{n-1}$.
\end{theorem}

@@

Combining all these facts together, we see that the number of permutations $\sigma$ in $T_{2m}$ that avoid $231$ must be equal to $ 2^{m-1}$, since every even element can be inserted in exactly two positions in $\sigma$.

@@

\begin{theorem}
$t_n(213)$, the number of permutations in $T_n$ avoiding the pattern 213, is equal to $2^{n-1}$.
\end{theorem}

@@

First, note that $2m-1, 2m$ must be inserted to the right of  $2m-3$, because of the way $T_{2m}$ is defined. Furthermore, if we insert $2m-1, 2m$ right after some $a$ in $\sigma$ not equal to $2m-3$ or $2m-2$, then we see that the produced permutation $\sigma$ has a three-element sub-sequence in the relative ordering $213$ -- $(2m-2, a, 2m-1)$. Hence we conclude that $2m-1, 2m$ must be inserted to the right of either $2m-3$ or $2m-2$; since we can do this for every permutation $\sigma'$ in $A_{m-1}$, we conclude that $|A_m| = 2|A_{m-1}|$. By the induction hypothesis, we have that $|A_{m-1}|=2^{m-2}$: so $A_m$ has size $2^{m-1}$, as desired, and the result follows by induction.
\end{proof}

>>>>

In Sections \ref{312} and \ref{321} we consider sequences $s_n(\pi)$ where $\pi$ is an element of  \\ $\{123,132,213,231,312,321\}$, and prove the following theorem.

\begin{theorem*}
For $\pi$ in $\{123,132,213,231,312,321\}$, the sequence $(s_n(\pi))$ is the Catalan numbers.
\end{theorem*}

Of particular interest is the fact that two entirely distinct methods of proof had to be used to prove this theorem for all six permutations. In Section \ref{S4} we will consider sequences $s_n(\pi)$ where $\pi$ is a permutation of $\{1,2,3,4\}$ and state a conjecture concerning $(s_n(\pi))$. In Section \ref{Tn} we consider a variation of the avoidance problem that considers the problem of avoiding sequences in a subset $T_n$ of the permutations of $n$ elements. Before proceeding, however, we formally introduce the notation that will be used throughout this paper.

@@

\emph{Yajit wrote this section. Leon closely edited this section. Deepak made minor edits to this section.}

@@

\subsection{Reversing}

We notate the reversing operator as $\mathcal{R}$. Formally, if $a_1 a_2 \ldots a_n$ is a permutation of $n$ elements, then $\mathcal{R}(a_1 a_2 \ldots a_{n-1} a_n) = a_n a_{n-1} \dots a_2 a_1$.

\begin{lemma}[Reversing Lemma]
The permutation $\sigma$ avoids the permutation $\pi$ if and only if $\mathcal{R}(\sigma)$ avoids $\mathcal{R}(\pi)$. I.e. $\mathcal{R}$ is an avoidance preserving operator. 
\end{lemma}
\begin{proof}
We need to only prove one direction of this theorem, then the other direction follows because $\mathcal{R}^2=1$. We show that if ${\sigma}\in S_n$ avoids $\pi$ then $\mathcal{R}(\sigma)$ avoids $\mathcal{R}(\pi)$. Suppose that $\mathcal{R}(\sigma)$ does not avoid $\mathcal{R}(\pi)$. Then if we reverse $\mathcal{R}(\sigma)$ again, the subsequence of $\mathcal{R}(\sigma)$ that follows the same pattern as $\mathcal{R}(\pi)$ will reverse to a subsequence that follows the same pattern as $\pi$, giving us a contradiction. Therefore $\mathcal{R}(\sigma)$ avoids $\mathcal{R}(\pi)$.

@@

\begin{corollary} 
For a permutation $\pi$, $s_n(\pi)=s_n(\mathcal{R}(\pi))$. 
\end{corollary}
\begin{proof}
This result follows trivially because the reversing lemma describes a bijection between permutations that avoid $\pi$ and permutations that avoid $\mathcal{R}(\pi)$. 
\end{proof}

\subsection{Flipping}
In this section we introduce the concept of flipping to prove the `Flipping Lemma'.
\begin{definition}
We define the \emph{flip} of a sequence $a$ as the sequence $b$ with the same elements as $a$, but with the largest element swapped with the smallest element, the second largest element swapped with the second smallest element, etc.  The flipping operator will be denoted by $\mathcal{F}$.  
\end{definition}

@@

\begin{lemma}[Flipping Lemma]
The permutation $\sigma$ avoids the permutation $\pi$ if and only if $\mathcal{F}(\sigma)$ avoids $\mathcal{F}(\pi)$. I.e. $\mathcal{F}$ is an avoidance preserving operator. 
\end{lemma}
\begin{proof}
Notice that we only need to prove one direction of this statement. The opposite direction will follow since applying $\mathcal{F}$ twice returns the original permutation, or $\mathcal{F}^2 = 1$. We show that if ${\sigma}\in S_n$ avoids $\pi$ then $\mathcal{F}(\sigma)$ avoids $\mathcal{F}(\pi)$. Suppose that $\mathcal{F}(\sigma)$ does not avoid $\mathcal{F}(\pi)$. Let $\mathcal{F}(\sigma)=a_1\cdots a_n$ and let $\mathcal{F}(\pi)=b_1\cdots  b_k$. Then there is some subsequence $a_{i_1}\cdots a_{i_k}$ so that $a_{i_c}<a_{i_d}$ if and only if $b_c<b_d$. Now flipping $\mathcal{F}(\sigma)$ gives us the sequence $(n+1-a_1)(n+1-a_2)\cdots(n+1-a_n)$ and flipping $\mathcal{F}(\pi)$ gives the sequence $(k+1-b_1)(k+1-b_2)\cdots(k+1-b_k)$. Thus $n+1-a_{i_c}>n+1-a_{i_d}$ if and only if $k+1-b_c>k+1-b_d$. So  this implies that $\sigma$ does not avoid $\pi$, a contradiction. 
\end{proof}

\begin{corollary} 
For a permutation $\pi$, $s_n(\pi)=s_n(\mathcal{F}(\pi))$. 
\end{corollary}
\begin{proof}
This result follows because the flipping lemma describes a bijection between permutations that avoid $\pi$ and permutations that avoid $\mathcal{F}(\pi)$. 
\end{proof}

\emph{Yajit wrote this section. Leon edited this section. Deepak made minor edits.}

@@

\emph{Deepak wrote this section. Leon closely edited this section. Yajit made minor revisions.}

@@

\label{321}
Unfortunately, there is no clear way to reformulate Lemma \ref{lemma1} for the case of avoiding the permutation 321 or 123. As a consequence, the proof in the previous section simply does not hold. Fortunately, however, we still have a closed form expression for $s_n(321)$ (recall that $s_n(321)=s_n(123)$ by our flipping lemma). Surprisingly, in fact, the sequence remains the same: $s_n(321)=C_n$, the $n^{th}$ Catalan number. In order to prove this result, we need the machinery of Young tableaux. What follows is a brief exposition of the key results, adapted from Chapter 8 of Stanley's \emph{Algebraic Combinatorics}.

@@

\emph{Leon wrote this section. Yajit made minor revisions.}

@@

Since $T_n$ is only defined for even $n$, wherever convenient, we shall refer to $T_n$ as $T_{2m}$ where $m$ is any integer greater than or equal to $1$.

@@

Unfortunately, it turns out that proving the equivalence between $231$ and $213$ is not as straightforward as above. In fact, however, we can prove a stronger result: both $t_n(231)$ and $t_n(213)$ are equal to $2^{n/2-1}$, and hence the sequences we obtain are equivalent.

\begin{theorem}
$t_n(231)$, the number of permutations in $T_n$ avoiding the pattern 231, where $n$ is even, is equal to $2^{n/2-1}$.
\end{theorem}

@@

Combining all these facts together, we see that the number of permutations $\sigma$ in $T_{2m}$ that avoid $231$ must be equal to $ 2^{m-1} = 2^{n/2-1}$, since every even element except $2m$ can be inserted in exactly two positions in $\sigma$.

@@

\begin{theorem}
$t_n(213)$, the number of permutations in $T_n$ avoiding the pattern 213, where $n$ is even, is equal to $2^{n/2-1}$.
\end{theorem}

@@

First, note that $2m-1, 2m$ must be inserted to the right of  $2m-3$, because of the way $T_{2m}$ is defined. Furthermore, if we insert $2m-1, 2m$ right after some $a$ in $\sigma$ not equal to $2m-3$ or $2m-2$, then we see that the produced permutation $\sigma$ has a three-element sub-sequence in the relative ordering $213$ -- $(2m-2, a, 2m-1)$. Hence we conclude that $2m-1, 2m$ must be inserted to the right of either $2m-3$ or $2m-2$; since we can do this for every permutation $\sigma'$ in $A_{m-1}$, we conclude that $|A_m| = 2|A_{m-1}|$. By the induction hypothesis, we have that $|A_{m-1}|=2^{m-2}$: so $A_m$ has size $2^{m-1} = 2^{n/2 - 1}$, as desired, and the result follows by induction.
\end{proof}
